---
title: "Milwaukee Bucks Shot Zone Analysis"
subtitle: "2022-23 NBA Season"
author: "Analytics Intern"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

```{r load-data, echo=FALSE}
# Purpose: Load required libraries and raw shot data
# Input:   bucks2023.csv — shot-level data for the 2022-23 Milwaukee Bucks season
# Output:  raw 'data' dataframe for downstream processing

library(tidyverse)
library(ggforce)

data <- read_csv("bucks2023.csv")
```



```{r circle-helper, echo=FALSE}
# Purpose: Generate x/y coordinates along an arc or full circle,
#          used to draw curved court elements and zone boundaries
# Inputs:
#   center_x, center_y — center point of the circle
#   radius             — radius of the circle
#   start, end         — start and end angles in radians (default: full circle)
#   n                  — number of points to generate (default: 500 for smooth curves)
# Output: dataframe with columns x and y

circle_points <- function(center_x, center_y, radius,
                          start = 0, end = 2 * pi, n = 500) {
  angles <- seq(start, end, length.out = n)
  data.frame(
    x = center_x + radius * cos(angles),
    y = center_y + radius * sin(angles)
  )
}
```

```{r court-geometry, echo=FALSE}
# Purpose: Define all NBA court elements as dataframes for use in ggplot
#
# Coordinate system notes:
#   - locationX ranges from -250 to 250 (tenths of feet); origin at the basket
#   - locationY is measured from the basket upward toward half court
#   - All court measurements are converted to tenths of feet to match this system
#   - The basket sits 5.25 ft (52.5 units) from the baseline, so the basket
#     is centered at (0, 52.5) in our plotting coordinate system
#   - Shot coordinates are shifted up by 52.5 units (see Section 4) so that
#     the baseline sits at y = 0, making court drawing more intuitive

# Angle where the 3PT arc meets the straight corner lines (at x = ±220, i.e. 22 ft)
cutoff_angle <- acos(220 / 237.5)

# Y value where the arc meets x = 220 — defines the corner line upper endpoint
corner_y <- sqrt(237.5^2 - 220^2) + 52.5

# Three-point arc: traces from right corner cutoff to left corner cutoff
three_pt <- circle_points(0, 52.5, 237.5,
                          start = cutoff_angle,
                          end   = pi - cutoff_angle)

# Free throw circle: centered at the free throw line (19 ft = 190 units from baseline)
ft_circle <- circle_points(0, 190, 60)

# Restricted area: 4 ft (40 unit) radius semi-circle above the basket
restricted <- circle_points(0, 52.5, 40, start = 0, end = pi)

# Basket: 9-inch radius (7.5 units) circle centered on the basket
basket <- circle_points(0, 52.5, 7.5)

# Center circle at half court (47 ft = 470 units from baseline)
center_circle <- circle_points(0, 470, 60)
```

---

# 1. Shot Chart — Individual Shots

```{r shot-chart, fig.width = 9, fig.height = 10, fig.align = "center", echo=FALSE}
# Purpose: Plot every shot attempt for the 2022-23 season on a half court diagram
# Input:   raw 'data' dataframe
# Output:  ggplot shot chart with made/missed color encoding
#
# Key decisions:
#   - locationY is shifted up by 52.5 units inline (+ 52.5) so shots align with
#     the court drawing, where the baseline is at y = 0 and the basket at y = 52.5
#   - Alpha = 0.3 and size = 0.8 used to handle overplotting across ~6,600 shots
#   - Bucks green (#00471B) for made shots; red (#CC1122) for missed shots
#   - Backcourt shots fall outside ylim and are not shown;
#     the caption notes their exclusion for transparency

ggplot() +

  # --- Court elements ---
  geom_rect(aes(xmin = -80, xmax = 80, ymin = 0, ymax = 190),
            fill = NA, color = "black") +
  geom_path(data = three_pt,      aes(x, y), color = "black") +
  geom_segment(aes(x =  220, xend =  220, y = 0, yend = corner_y), color = "black") +
  geom_segment(aes(x = -220, xend = -220, y = 0, yend = corner_y), color = "black") +
  geom_segment(aes(x = -250, xend =  250, y = 0,   yend = 0),   color = "black") +
  geom_segment(aes(x = -250, xend =  250, y = 470, yend = 470), color = "black") +
  geom_path(data = ft_circle,     aes(x, y), color = "black") +
  geom_path(data = restricted,    aes(x, y), color = "black") +
  geom_path(data = basket,        aes(x, y), color = "black") +
  geom_path(data = center_circle, aes(x, y), color = "black") +

  # --- Shot points ---
  geom_point(data = data,
             aes(x = locationX, y = locationY + 52.5, color = isShotMade),
             alpha = 0.3, size = 0.8) +

  # --- Formatting ---
  scale_color_manual(values = c("TRUE"  = "#00471B", "FALSE" = "#CC1122"),
                     labels  = c("TRUE"  = "Made",   "FALSE" = "Missed")) +
  coord_fixed() +
  xlim(-260, 260) +
  ylim(-10, 520) +
  theme_minimal() +
  theme(axis.text  = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  labs(title    = "Milwaukee Bucks Shot Chart",
       subtitle = "2022-23 NBA Season",
       color    = "Shot Result",
       x        = NULL,
       y        = NULL,
       caption  = "Backcourt shots excluded")
```
```{r zone-assignment, echo=FALSE}
# Purpose: Assign each shot to one of 9 custom shooting zones
# Input:   raw 'data' dataframe
# Output:  'data' with new 'custom_zone' column
#
# The 9 zones are:
#   Paint, Mid Range Right, Mid Range Left, Mid Range Top,
#   Right Corner 3, Left Corner 3, Right Wing 3, Left Wing 3, Top of Key 3
#
# Key decisions:
#   - case_when evaluates in order; Paint is checked first to prevent
#     mid-range conditions from capturing shots inside the key
#   - distanceShot is in feet; 3PT line radius = 23.75 ft used for two/three split
#   - Zone boundaries (x = ±80, ±100, ±220, y = 150, 190) were chosen to match
#     standard NBA court markings and align with the polygon zones drawn below

data <- data %>%
  mutate(custom_zone = case_when(
    abs(locationX) <= 80 & locationY <= 190                    ~ "Paint",
    locationX >  0 & locationY <= 190 & distanceShot < 23.75  ~ "Mid Range Right",
    locationX <  0 & locationY <= 190 & distanceShot < 23.75  ~ "Mid Range Left",
    locationY > 190 & distanceShot < 23.75                    ~ "Mid Range Top",
    locationX >  100 & locationY <= 150                        ~ "Right Corner 3",
    locationX < -100 & locationY <= 150                        ~ "Left Corner 3",
    locationX >  100 & locationY >  150                        ~ "Right Wing 3",
    locationX < -100 & locationY >  150                        ~ "Left Wing 3",
    abs(locationX) <= 100                                      ~ "Top of Key 3",
    TRUE                                                       ~ "Other"
  ))
```

---

# 2. Zone Shooting Statistics

```{r zone-stats, echo=FALSE}
# Purpose: Calculate shooting percentage and mean shot location per zone
# Input:   'data' with custom_zone
# Output:  custom_zone_stats — one row per zone with attempts, makes, shoot_pct,
#          and mean x/y coordinates used for label placement on the chart
#
# Judgment: Backcourt shots (locationY > 470 + 52.5 = 522.5 units) are excluded
#           as they are low-percentage desperation heaves and not meaningful
#           shot selection decisions. This is noted in the chart caption.

custom_zone_stats <- data %>%
  filter(locationY <= 522.5) %>%
  group_by(custom_zone) %>%
  summarise(
    attempts  = n(),
    makes     = sum(isShotMade),
    shoot_pct = makes / attempts,
    mean_x    = mean(locationX),
    mean_y    = mean(locationY)
  )

custom_zone_stats
```
```{r zone-polygons, echo=FALSE}
# Purpose: Define each of the 9 shooting zones as a polygon for geom_polygon
#          Zones follow the 3PT arc where applicable rather than using rectangles,
#          ensuring zones do not overlap or breach court boundaries
# Output:  one dataframe per zone, combined into 'all_zones' with shoot_pct joined

# --- Key angle and boundary calculations ---

# Angle where arc crosses x = ±100 (wing / top-of-key boundary)
wing_angle <- acos(100 / 237.5)

# Angle where arc crosses y = 150 (corner / wing boundary); adjusted for basket offset
corner_angle <- asin((150 - 52.5) / 237.5)

# Angles where arc crosses y = 190 (free throw line); used to cap mid-range zones
mid_right_angle_outer <- acos(220 / 237.5)
mid_right_angle_top   <- asin((190 - 52.5) / 237.5)
mid_top_angle_right   <- mid_right_angle_top
mid_top_angle_left    <- pi - mid_top_angle_right

# --- Arc segments ---

# Mid-range right arc: from y = 190 down to where arc meets x = 220
mid_right_arc <- circle_points(0, 52.5, 237.5,
                                start = mid_right_angle_top,
                                end   = mid_right_angle_outer)

# Mid-range left arc: mirror of right
mid_left_arc <- circle_points(0, 52.5, 237.5,
                               start = pi - mid_right_angle_top,
                               end   = pi - mid_right_angle_outer)

# Mid-range top arc: spans full width of mid-range at y = 190
mid_top_arc <- circle_points(0, 52.5, 237.5,
                              start = mid_top_angle_right,
                              end   = mid_top_angle_left)

# Right wing arc: from corner boundary up to wing boundary
right_wing_arc <- circle_points(0, 52.5, 237.5,
                                 start = corner_angle,
                                 end   = wing_angle)

# Left wing arc: mirror of right
left_wing_arc <- circle_points(0, 52.5, 237.5,
                                start = pi - wing_angle,
                                end   = pi - corner_angle)

# Top of key arc: from right wing boundary to left wing boundary
top_key_arc <- circle_points(0, 52.5, 237.5,
                              start = wing_angle,
                              end   = pi - wing_angle)

# --- Zone polygons ---
# Points are ordered sequentially to trace each polygon without self-intersection
# filter(y <= 190) applied to mid-range side zones to clip floating point overshoot

paint_zone <- data.frame(
  x = c(-80, 80, 80, -80),
  y = c(0, 0, 190, 190)
)

mid_right_zone <- rbind(
  data.frame(x = 80,  y = 0),
  data.frame(x = 220, y = 0),
  data.frame(x = 220, y = 141.9777),
  data.frame(x = rev(mid_right_arc$x), y = rev(mid_right_arc$y)),
  data.frame(x = 80,  y = 190)
) %>% filter(y <= 190)

mid_left_zone <- rbind(
  data.frame(x = -220, y = 0),
  data.frame(x = -80,  y = 0),
  data.frame(x = -80,  y = 190),
  data.frame(x = mid_left_arc$x, y = mid_left_arc$y),
  data.frame(x = -220, y = 141.9777)
) %>% filter(y <= 190)

mid_top_zone <- rbind(
  data.frame(x = 220,  y = 190),
  data.frame(x = mid_top_arc$x, y = mid_top_arc$y),
  data.frame(x = -220, y = 190)
)

right_corner_zone <- data.frame(
  x = c(220, 250, 250, 220),
  y = c(0, 0, 150, 150)
)

left_corner_zone <- data.frame(
  x = c(-250, -220, -220, -250),
  y = c(0, 0, 150, 150)
)

right_wing_zone <- rbind(
  data.frame(x = 220, y = 0),
  data.frame(x = 250, y = 0),
  data.frame(x = 250, y = 470),
  data.frame(x = 100, y = 470),
  data.frame(x = rev(right_wing_arc$x), y = rev(right_wing_arc$y))
)

left_wing_zone <- rbind(
  data.frame(x = -250, y = 0),
  data.frame(x = -220, y = 0),
  data.frame(x = rev(left_wing_arc$x), y = rev(left_wing_arc$y)),
  data.frame(x = -100, y = 470),
  data.frame(x = -250, y = 470)
)

top_key_zone <- rbind(
  data.frame(x = -100, y = 470),
  data.frame(x =  100, y = 470),
  data.frame(x = top_key_arc$x, y = top_key_arc$y)
)

# --- Combine all zones and join shooting percentage ---
all_zones <- rbind(
  cbind(paint_zone,        zone = "Paint"),
  cbind(mid_right_zone,    zone = "Mid Range Right"),
  cbind(mid_left_zone,     zone = "Mid Range Left"),
  cbind(mid_top_zone,      zone = "Mid Range Top"),
  cbind(right_corner_zone, zone = "Right Corner 3"),
  cbind(left_corner_zone,  zone = "Left Corner 3"),
  cbind(right_wing_zone,   zone = "Right Wing 3"),
  cbind(left_wing_zone,    zone = "Left Wing 3"),
  cbind(top_key_zone,      zone = "Top of Key 3")
) %>%
  left_join(custom_zone_stats %>% select(custom_zone, shoot_pct),
            by = c("zone" = "custom_zone"))
```

---

# 3. Zone Efficiency Chart

```{r zone-chart, fig.width = 9, fig.height = 10, fig.align = "center", echo=FALSE}
# Purpose: Visualize shooting efficiency by zone on a half court diagram
# Input:   all_zones (polygons + shoot_pct), custom_zone_stats (label positions),
#          court element dataframes from Section 3
# Output:  ggplot zone efficiency chart with color-coded zones and % labels
#
# Design decisions:
#   - Zones drawn first, court lines on top so markings remain visible
#   - Color gradient: red (low) → neon green (mid) → dark Bucks green (high)
#   - Midpoint set to 0.44 to approximate league average FG%, so below-average
#     zones appear red and above-average zones appear green
#   - Limits set to actual data range (0.32–0.60) for maximum color contrast
#   - Labels offset +40 units on y-axis to avoid overlap with court markings

ggplot() +

  # --- Zone polygons (drawn first, underneath court lines) ---
  geom_polygon(data = all_zones,
               aes(x = x, y = y, group = zone, fill = shoot_pct),
               alpha = 0.6, color = "black") +

  scale_fill_gradient2(
    low      = "#FF3131",
    mid      = "#39FF14",
    high     = "#00471B",
    midpoint = 0.44,
    limits   = c(0.32, 0.60),
    labels   = scales::percent,
    name     = "Shooting %"
  ) +

  # --- Court lines (drawn on top of zones) ---
  geom_rect(aes(xmin = -80, xmax = 80, ymin = 0, ymax = 190),
            fill = NA, color = "black") +
  geom_path(data = three_pt,      aes(x, y), color = "black") +
  geom_segment(aes(x =  220, xend =  220, y = 0, yend = corner_y), color = "black") +
  geom_segment(aes(x = -220, xend = -220, y = 0, yend = corner_y), color = "black") +
  geom_segment(aes(x = -250, xend =  250, y = 0,   yend = 0),   color = "black") +
  geom_segment(aes(x = -250, xend =  250, y = 470, yend = 470), color = "black") +
  geom_path(data = ft_circle,     aes(x, y), color = "black") +
  geom_path(data = restricted,    aes(x, y), color = "black") +
  geom_path(data = basket,        aes(x, y), color = "black") +
  geom_path(data = center_circle, aes(x, y), color = "black") +

  # --- Zone shooting % labels ---
  geom_label(data = custom_zone_stats,
             aes(x = mean_x, y = mean_y + 40,
                 label = scales::percent(shoot_pct, accuracy = 0.1)),
             size = 3, alpha = 0.8) +

  # --- Formatting ---
  coord_fixed() +
  xlim(-260, 260) +
  ylim(-10, 520) +
  theme_minimal() +
  theme(axis.text  = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  labs(title    = "Milwaukee Bucks Shot Zone Efficiency",
       subtitle = "2022-23 NBA Season",
       fill     = "Shooting %",
       x        = NULL,
       y        = NULL,
       caption  = "Backcourt shots excluded")
```

---

# 4. Key Findings

## Summary

- **Paint dominates efficiency:** The Bucks convert nearly 59% of shots in the restricted area/paint, more than any other zone
- **Corner three disparity:** The left corner shooting percentage (44%) is way higher than the right corner (32%)
- **Mid-range is consistent:** All three mid-range zones sit between 39-42%, just below league average
- **Wing threes underperform:** Both wings sit around 35-37%, suggesting the team should seek higher quality three point looks

---

## Detailed Findings

### Paint Efficiency (59.1%, 3,334 attempts)
The paint is by far the most efficient zone at 59.1%. This suggests the Bucks are effective at generating high-percentage looks near the basket, whether through drives, post-ups, or offensive rebounds.I would assume this is largely because of Giannis. A dominating paint presence.

### Left vs. Right Corner Three (44.1% vs. 32.2%)
This 12-point gap is the most striking finding in the data. Corner threes are generally considered the most efficient three-point shots in basketball due to the shorter distance (22 ft vs. 23.75 ft). The Bucks are capitalizing on this on the left side but not the right. This asks the question — which players are taking right corner threes, and are defensive rotations forcing lower-quality looks to that spot?

### Mid-Range Zones (39–42%)
Mid-range shots are widely regarded as the least efficient shot type in modern basketball — too far for a high percentage, too close to count for three points. The Bucks' 39–42% from these zones is roughly average. The very low volume from mid-range top (60 attempts) suggests the team is already limiting these looks, which is consistent with modern offensive strategy.

### Wing Threes (35.4% right, 37.5% left)
Wing threes account for a combined 1,477 attempts — a significant portion of the shot profile. At 35–37%, these are below the efficiency threshold many teams target (~38%+) for high-volume three-point zones. Improving shot quality in these zones (e.g., more open looks off movement) or shifting volume toward the corners could improve overall offensive efficiency.

### Top of Key Threes (36.3%, 835 attempts)
The top of key is the Bucks' highest-volume three-point zone and also their least efficient. At 36.3%, it is marginally above break-even efficiency but represents an area where shot selection could be improved — particularly by identifying which players are taking these shots and under what defensive conditions.

---

# 4 Point Line Suggestions

```{r, echo=FALSE}
# 28 ft + 5.25 ft basket offset = 33.25 ft = 332.5 units from baseline
# But the arc radius is measured from the basket, so just 28 ft = 280 units radius
four_pt_arc <- circle_points(0, 52.5, 280, start = cutoff_angle, end = pi - cutoff_angle)

# Corner straight lines for 4pt arc
four_pt_corner_y <- sqrt(280^2 - 220^2) + 52.5


ggplot() +
  # Paint
  geom_rect(aes(xmin = -80, xmax = 80, ymin = 0, ymax = 190),
            fill = NA, color = "black") +
  
  # Three point arc
  geom_path(data = three_pt, aes(x, y), color = "black") +
  
  # Corner three lines
  geom_segment(aes(x = 220, xend = 220, y = 0, yend = corner_y), color = "black") +
  geom_segment(aes(x = -220, xend = -220, y = 0, yend = corner_y), color = "black") +
  
  # Baseline
  geom_segment(aes(x = -250, xend = 250, y = 0, yend = 0), color = "black") +
  
  # Half court line
  geom_segment(aes(x = -250, xend = 250, y = 470, yend = 470), color = "black") +
  
  # Free throw circle
  geom_path(data = ft_circle, aes(x, y), color = "black") +
  
  # Restricted area
  geom_path(data = restricted, aes(x, y), color = "black") +
  
  # Basket
  geom_path(data = basket, aes(x, y), color = "black") +
  
  # Center circle
  geom_path(data = center_circle, aes(x, y), color = "black") +
  
  # Four point arc
  geom_path(data = four_pt_arc, aes(x, y), color = "blue", linewidth = 1, linetype = "dashed") +
  
  # Four point corner lines
  geom_segment(aes(x =  220, xend =  220, y = corner_y, yend = four_pt_corner_y), 
               color = "blue", linewidth = 1, linetype = "dashed") +
  geom_segment(aes(x = -220, xend = -220, y = corner_y, yend = four_pt_corner_y), 
               color = "blue", linewidth = 1, linetype = "dashed")

```

## Where to draw the 4-point line:
Based on expected value calculations, a 4-point line at 28 feet is the best placement. At 28 feet, the Bucks shot 33.7%, producing an expected value of 1.35 points per attempt — still exceeding the expected value of an average 3-point attempt (~1.25). At 29 feet, EV drops to 0.88, making any placement beyond 28 feet unjustified.

## Arc shape:
The 4-point arc follows the same shape as the existing 3-point arc — a circular arc with a radius of 28 feet.

## Consequence:
Grayson Allen would be the biggest beneficiary on the 2022-23 Bucks roster. He made 19 shots from beyond 28 feet, which currently count for 57 total points. Under a 4-point rule, those same makes would be worth 76 points — an increase of approximately 3% in his total season point output. While not much, this would meaningfully reward players who are good at shooting from deep and could change how teams construct rosters.